A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-04-15 06-appendix.tex
Add artifact evaluation
Cleanup and compress stuff in Section 3 up to Determinism

\clearpage
\appendix
\section{Appendix: Detailed Proofs (optional)}
\label{sec.proofs}

\subsection*{Determinism}

\begin{lemma}
  \label{lem.det-out}
  %%
  If~$\delta\out\delta_1$ and~$\delta\out\delta_2$ then~$\delta_1=\delta_2$.
\end{lemma}
\begin{proof}
  The lemma is vacuously true if~$\delta$ cannot be advanced by~$\out$
  transitions.  Suppose that is not the case and let~$\delta=\<p,n,e>$,
  $\delta_1=\<p_1,n_1,e_1>$ and~$\delta_2=\<p_2,n_2,e_2>$.  Then, there are
  two possibilities.
  \begin{case}
    $e\ne\nil$.  Both transitions are applications of~\R{push}.
    Hence~$p_1=p_2=\bcast(p,e)$, $n_1=n_2=n+1$, and~$e_1=e_2=\nil$.
  \end{case}
  \begin{case}
    $e=\nil$.  Both transitions are applications of~\R{pop}.
    Hence~$p_1=p_2=p$, $n_1=n_2=n-1$, and~$e_1=e_2=\nil$.\qedhere
  \end{case}
\end{proof}

% \begin{theorem}
%   \label{thm.det-out-i}
%   %%
%   If~$\delta\out[i]\delta_1$ and~$\delta\out[i]\delta_2$
%   then~$\delta_1=\delta_2$.
% \end{theorem}
% \begin{proof}
%   By induction on~$i$.
%   %%
%   The theorem is trivially true if~$i=0$ and follows directly from
%   Lemma~\ref{lem.det-out} if~$i=1$.  (The theorem is vacuously true
%   for~$\outpush$ transitions if~$i>1$ since they cannot be applied more
%   than once in a row and cannot occur after a~$\outpop$ transition.)
%   %%
%   Suppose
%   \[
%     \delta\out[1]\delta_1'\out[i-1]\delta_1
%     \quad\text{and}\quad
%     \delta\out[1]\delta_2'\out[i-1]\delta_2\,,
%   \]
%   for some~$i>1$, $\delta_1'$ and~$\delta_2'$.
%   %%
%   By Lemma~\ref{lem.det-out}, $\delta_1'=\delta_2'$.  By the induction
%   hypothesis, $\delta_1=\delta_2$.\qedhere
% \end{proof}

\begin{lemma}
  \label{lem.det-nst}
  %%
  If~$\delta\nst\delta_1$ and~$\delta\nst\delta_2$ then~$\delta_1=\delta_2$.
\end{lemma}
\begin{proof}
  By induction on the structure of~$\nst$ derivations.  The lemma is
  vacuously true if~$\delta$ cannot be advanced by~$\nst$ transitions.
  Suppose that is not the case and let~$\delta=\<p,n,e>$,
  $\delta_1=\<p_1,n_1,e_1>$ and~$\delta_2=\<p_2,n_2,e_2>$.  Then, by the
  hypothesis of the lemma, there are derivations~$\pi_1$ and~$\pi_2$ such
  that
  \begin{align*}
    \pi_1&\Vdash\<p,n,e>\nst\<p_1,n_1,e_1>\\
    \pi_2&\Vdash\<p,n,e>\nst\<p_2,n_2,e_2>
  \end{align*}
  i.e., the conclusion of~$\pi_1$ is~$\<p,n,e>\nst\<p_1,n_1,e_1>$ and the
  conclusion of~$\pi_2$ is~$\<p,n,e>\nst\<p_2,n_2,e_2>$.

@@

  Depending on the structure of program~$p$, the following~11 cases are
  possible.  (Note that~$p$ cannot be an~$\ceu{\AwaitExt}$,
  $\ceu{\AwaitInt}$, $\ceu{\Break}$, $\ceu{\Every}$, $\ceu{\Fin}$,
  or~$\ceu{\Nop}$ expression as there are no~$\nst$ rules to transition such
  programs.)

  \begin{case}
    $p=\ceu{\Mem(\Id)}$.
    %%
    Then derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{mem},
    i.e., their conclusions are obtained by an application of this rule.
    Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
  \end{case}

  \begin{case}
    $p=\ceu{\EmitInt(e')}$.
    %%
    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{emit-int}.
    Hence~$p_1=p_2=\ceu{\CanRun(n)}$ and~$e_1=e_2=e'$.
  \end{case}

  \begin{case}
    $p=\ceu{\CanRun(n)}$.
    %%
    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{can-run}.
    Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
  \end{case}

  \begin{case}
    $p=\ceu{\IfElse{\Mem(\Id)}{p'}{p''}}$.
    %%
    There are two subcases.
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$~is true.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-true}.
      Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$ is false.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-false}.
      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{p';\,p''}$.
    %%
    There are three subcases.
    \begin{subcase}
      $p'=\ceu{\Nop}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-nop}.
      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'=\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-brk}.
      Hence~$p_1=p_2=\ceu{\Break}$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-adv}.
      Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
      \begin{align*}
        \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
        \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
      \end{align*}
      for some~$p_1'$, $p_2'$, $e_1'$, and~$e_2'$.  By the induction
      hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
      Hence~$p_1=\ceu{p_1';p''}=\ceu{p_2';p''}=p_2$ and~$e_1=e_1'=e_2'=e_2$.
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{\Loop{p'}}$.
    %%
    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-expd}.
    Hence~$p_1=p_2=\ceu{p'\AtLoop{p'}}$ and~$e_1=e_2=\nil$.
  \end{case}

  \begin{case}
    $p=\ceu{p'\AtLoop{p''}}$.
    %%
    There are three subcases.
    \begin{subcase}
      $p'=\ceu{\Nop}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-nop}.
      Hence~$p_1=p_2=\ceu{\Loop{p''}}$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'=\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-brk}.
      Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-adv}.
      Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
      \begin{align*}
        \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
        \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
      \end{align*}
      for some~$p_1'$, $p_2'$, $e_1'$, and~$e_2'$.  By the induction
      hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
      Hence~$p_1=\ceu{p_1'\AtLoop{p''}}=\ceu{p_2'\AtLoop{p''}}=p_2$
      and~$e_1=e_1'=e_2'=e_2$.
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{p'\And{p''}}$.
    %%
    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-expd}.
    Hence~$p_1=p_2=\ceu{{p'}\AtAnd{(\CanRun(n);\,p'')}}$ and~$e_1=e_2=\nil$.
  \end{case}

  \begin{case}
    $p=\ceu{p'\AtAnd{p''}}$.
    %%
    There are two subcases.
    \begin{subcase}
      $\isblocked(p',n)$~is false.
      %%
      There are three subcases.
      \begin{subsubcase}
        $p'=\ceu{\Nop}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop1}.
        Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
      \end{subsubcase}
      \begin{subsubcase}
        \label{lem.det-nst.and-brk1}
        $p'=\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk1}.
        Hence~$p_1=p_2=\ceu{\clear(p'');\Break}$ and~$e_1=e_2=\nil$.
      \end{subsubcase}
      \begin{subsubcase}
        \label{lem.det-nst.and-adv1}
        $p'\ne\ceu{\Nop},\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv1}.
        Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
        \begin{align*}
          \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
          \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
        \end{align*}
        for some~$p_1'$, $p_2'$, $e_1'$, $e_2'$.  By the induction
        hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
        Hence~$p_1=\ceu{{p_1'}\And{p''}}=\ceu{{p_2'}\And{p''}}=p_2$
        and~$e_1=e_1'=e_2'=e_2$.
      \end{subsubcase}
    \end{subcase}
    \begin{subcase}
      $\isblocked(p',n)$~is true.
      %%
      There are three subcases.
      \begin{subsubcase}
        $p''=\ceu{\Nop}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop2}.
        Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
      \end{subsubcase}
      \begin{subsubcase}
        \label{lem.det-nst.and-brk2}
        $p''=\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk2}.
        Hence~$p_1=p_2=\ceu{\clear(p');\Break}$ and~$e_1=e_2=\nil$.
      \end{subsubcase}
      \begin{subsubcase}
        \label{lem.det-nst.and-adv2}
        $p''\ne\ceu{\Nop},\ceu{\Break}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv2}.  Thus there
        are derivations~$\pi_1''$ and~$\pi_2''$ such that
        \begin{align*}
          \pi_1''&\Vdash\<p'',n,\nil>\nst\<p_1'',n,e_1''>\\
          \pi_2''&\Vdash\<p'',n,\nil>\nst\<p_2'',n,e_2''>
        \end{align*}
        for some~$p_1''$, $p_2''$, $e_1''$, and~$e_2''$.  By the induction
        hypothesis, $p_1''=p_2''$ and~$e_1''=e_2''$.
        Hence~$p_1=\ceu{{p'}\And{p_1''}}=\ceu{{p'}\And{p_2''}}=p_2$
        and~$e_1=e_1''=e_2''=e_2$.
      \end{subsubcase}
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{p'\Or{p''}}$.
    %%
    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-expd}.
    Hence~$p_1=p_2=\ceu{{p'}\AtOr{(\CanRun(n);\,p'')}}$ and~$e_1=e_2=\nil$.
  \end{case}

  \begin{case}
    $p=\ceu{p'\AtOr{p''}}$.
    %%
    There are two subcases.
    \begin{subcase}
      $\isblocked(p',n)$~is false.
      %%
      There are three subcases.
      \begin{subsubcase}
        $p'=\ceu{\Nop}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-nop1}.
        Hence~$p_1=p_2=\clear(p'')$ and~$e_1=e_2=\nil$.
      \end{subsubcase}
      \begin{subsubcase}
        $p'=\ceu{\Break}$.
        %%
        Similar to Case~\ref{lem.det-nst.and-brk1}.
      \end{subsubcase}
      \begin{subsubcase}
        $p'\ne\ceu{\Nop},\ceu{\Break}$.
        %%
        Similar to Case~\ref{lem.det-nst.and-adv1}.
      \end{subsubcase}
    \end{subcase}
    \begin{subcase}
      $\isblocked(p',n)$~is true.
      %%
      There are three subcases.
      \begin{subsubcase}
        $p''=\ceu{\Nop}$.
        %%
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-nop1}.
        Hence~$p_1=p_2=\clear(p')$ and~$e_1=e_2=\nil$.
      \end{subsubcase}
      \begin{subsubcase}
        $p''=\ceu{\Break}$.
        %%
        Similar to Case~\ref{lem.det-nst.and-brk2}.
      \end{subsubcase}
      \begin{subsubcase}
        $p''\ne\ceu{\Nop},\ceu{\Break}$.
        %%
        Similar to Case~\ref{lem.det-nst.and-adv2}.
        %%
        \qedhere
      \end{subsubcase}
    \end{subcase}
  \end{case}
\end{proof}

% \begin{theorem}
%   \label{thm.det-nst-i}
%   %%
%   If~$\delta\nst[i]\delta_1$ and~$\delta\nst[i]\delta_2$
%   then~$\delta_1=\delta_2$.
% \end{theorem}
% \begin{proof}
%   By induction on~$i$.  Similar to the proof of
%   Theorem~\ref{thm.det-out-pop-i}.
% \end{proof}

\begin{theorem}[Determinism]\label{thm.det}\strut\\
  %%
  If~$\delta\trans[i]\delta_1$ and~$\delta\trans[i]\delta_2$
  then~$\delta_1=\delta_2$.
\end{theorem}
\begin{proof}
  By induction on~$i$.  The theorem is trivially true if~$i=0$ and follows
  directly from Lemmas~\ref{lem.det-out} and~\ref{lem.det-nst} for~$i=1$.
  %%
  Suppose
  \[
    \delta\trans[1]\delta_1'\trans[i-1]\delta_1
    \quad\text{and}\quad
    \delta\trans[1]\delta_2'\trans[i-1]\delta_2\,,
  \]
  for some~$i>1$, $\delta_1'$ and~$\delta_2'$.
  There are two possibilities.
  \begin{case}
    $\delta\out[1]\delta_1'$ and~$\delta\out[1]\delta_2'$.  Then, by
    Lemma~\ref{lem.det-out}, $\delta_1'=\delta_2'$, and by the induction
    hypothesis, $\delta_1=\delta_2$.
  \end{case}
  \begin{case}
    $\delta\nst[1]\delta_1'$ and~$\delta\nst[1]\delta_2'$.  Then, by
    Lemma~\ref{lem.det-nst}, $\delta_1'=\delta_2'$, and by the induction
    hypothesis, $\delta_1=\delta_2$.\qedhere
  \end{case}
\end{proof}

\subsection*{Termination}

@@

  \item Similar to Case~\eqref{lem.props-nst-i.e}.\qedhere
  \end{enumerate}
\end{proof}

\begin{assumption}[Syntactic restrictions]\strut
  \label{ass.syn-rest}
  %%
  \begin{enumerate}[(a)]
  \item\label{ass.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contain no
    occurrences of~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$,
    or~$\ceu{\Fin}$ expressions.  Consequently, for any~$n$,
    \[
      \<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>\,.
    \]
    %%
  \item\label{ass.syn-rest.loop} If~$p=\ceu{\Loop{p_1}}$ then all execution
    paths of~$p_1$ contain at least one occurrence of a
    matching~$\ceu{\Break}$ or an~$\ceu{\AwaitExt}$ expression.
    Consequently, for all~$n$, there are~$p_1'$ and~$e$ such that
    \[
      \<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>\,,
    \]
    where~$p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
  \end{enumerate}
\end{assumption}

@@

\begin{lemma}\strut
  \label{lem.rank-out}
  %%
  \begin{enumerate}[(a)]
  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
    then~$\rank(\delta)=\rank(\delta')$.
  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
    then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate}
\end{lemma}
\begin{proof}\strut
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$,
  $\rank(\delta)=\<i,j>$, and~$\rank(\delta')=\<i',j'>$.
  \begin{enumerate}[(a)]
  \item Suppose~$\<p,n,e>\outpush\<p',n',e'>$.  Then, by~\R{push},
    $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
    Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
    as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.
    %%
    It remains to be shown that~$i=i'$:
    \begin{align*}
      i&=\pot(p,e)
         \tag*{by Definition~\ref{def.rank}}\\
       &=\pot'(\bcast(p,e))
         \tag*{by Definition~\ref{def.pot}}\\
       &=\pot'(p')
         \tag*{since~$p'=\bcast(p,e)$}\\
       &=\pot'(\bcast(p',\nil))
         \tag*{by definition of~$\bcast$}\\
       &=\pot'(\bcast(p',e'))
         \tag*{since~$e'=\nil$}\\
       &=\pot(p',e')
         \tag*{by Definition~\ref{def.pot}}\\
       &=i'
         \tag*{by Definition~\ref{def.rank}}
    \end{align*}
    Therefore, $\<i,j>=\<i',j'>$.

>>>>

\section{Detailed Proofs}
\label{sec.proofs}

\lemxdetout*
\begin{proof}
  The lemma is vacuously true if~$\delta$ cannot be advanced by~$\out$
  transitions.  Suppose that is not the case and let~$\delta=\<p,n,e>$,
  $\delta_1=\<p_1,n_1,e_1>$ and~$\delta_2=\<p_2,n_2,e_2>$.  Then, there are
  two possibilities.
  \begin{casex}
  \item $e\ne\nil$.  Both transitions are applications of~\R{push}.
    Hence~$p_1=p_2=\bcast(p,e)$, $n_1=n_2=n+1$, and~$e_1=e_2=\nil$.
    %%
  \item $e=\nil$.  Both transitions are applications of~\R{pop}.
    Hence $p_1=p_2=p$, $n_1=n_2=n-1$, and~$e_1=e_2=\nil$.\qedhere
  \end{casex}
\end{proof}

\lemxdetnst*
\begin{proof}
  By induction on the structure of~$\nst$ derivations.  The lemma is
  vacuously true if~$\delta$ cannot be advanced by~$\nst$ transitions.
  Suppose that is not the case and let~$\delta=\<p,n,e>$,
  $\delta_1=\<p_1,n_1,e_1>$ and~$\delta_2=\<p_2,n_2,e_2>$.  Then, by the
  hypothesis of the lemma, there are derivations~$\pi_1$ and~$\pi_2$ such
  that
  \begin{align*}
    \pi_1&\Vdash\<p,n,e>\nst\<p_1,n_1,e_1>\\
    \pi_2&\Vdash\<p,n,e>\nst\<p_2,n_2,e_2>
  \end{align*}
  i.e., the conclusion of~$\pi_1$ is~$\<p,n,e>\nst\<p_1,n_1,e_1>$ and the
  conclusion of~$\pi_2$ is~$\<p,n,e>\nst\<p_2,n_2,e_2>$.

@@

  Depending on the structure of program~$p$, the following~11 cases are
  possible.  (Note that~$p$ cannot be an~$\ceu{\AwaitExt}$,
  $\ceu{\AwaitInt}$, $\ceu{\Break}$, $\ceu{\Every}$, $\ceu{\Fin}$,
  or~$\ceu{\Nop}$ statement as there are no~$\nst$ rules to transition such
  programs.)

  \begin{casex}
  \item $p=\ceu{\Mem(\Id)}$.
    Then derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{mem},
    i.e., their conclusions are obtained by an application of this rule.
    Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
  \item $p=\ceu{\EmitInt(e')}$.
    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{emit-int}.
    Hence~$p_1=p_2=\ceu{\CanRun(n)}$ and~$e_1=e_2=e'$.
  \item $p=\ceu{\CanRun(n)}$.
    %% \
    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{can-run}.
    Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
  \item $p=\ceu{\IfElse{\Mem(\Id)}{p'}{p''}}$.
    \begin{casex}
    \item$\eval(\ceu{\Mem(\Id)})$.
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-true}.
      Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
    \item$\lnot\eval(\ceu{\Mem(\Id)})$.
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-false}.
      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
    \end{casex}
  \item$p=\ceu{p';\,p''}$.
    \begin{casex}
    \item $p'=\ceu{\Nop}$.
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-nop}.
      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
    \item $p'=\ceu{\Break}$.
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-brk}.
      Hence~$p_1=p_2=\ceu{\Break}$ and~$e_1=e_2=\nil$.
    \item $p'\ne\ceu{\Nop},\ceu{\Break}$.
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-adv}.  Thus there
      are derivations~$\pi_1'$ and~$\pi_2'$ such that
        \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
        \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
      for some~$p_1'$, $p_2'$, $e_1'$, and~$e_2'$.  By the induction
      hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
      Hence~$p_1=\ceu{p_1';p''}=\ceu{p_2';p''}=p_2$
      and~$e_1=e_1'=e_2'=e_2$.
    \end{casex}
  \item $p=\ceu{\Loop{p'}}$.
    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-expd}.
    Hence~$p_1=p_2=\ceu{p'\AtLoop{p'}}$ and~$e_1=e_2=\nil$.
  \item $p=\ceu{p'\AtLoop{p''}}$.
    \begin{casex}
    \item $p'=\ceu{\Nop}$.
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-nop}.
      Hence~$p_1=p_2=\ceu{\Loop{p''}}$ and~$e_1=e_2=\nil$.
    \item $p'=\ceu{\Break}$.
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-brk}.
      Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
    \item $p'\ne\ceu{\Nop},\ceu{\Break}$.
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-adv}.  Thus there
      are derivations~$\pi_1'$ and~$\pi_2'$ such that
        \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
        \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
      for some~$p_1'$, $p_2'$, $e_1'$, and~$e_2'$.  By the induction
      hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
      Hence~$p_1=\ceu{p_1'\AtLoop{p''}}=\ceu{p_2'\AtLoop{p''}}=p_2$
      and~$e_1=e_1'=e_2'=e_2$.
    \end{casex}
  \item $p=\ceu{p'\And{p''}}$.
    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-expd}.
    Hence~$p_1=p_2=\ceu{{p'}\AtAnd{(\CanRun(n);\,p'')}}$
    and~$e_1=e_2=\nil$.
  \item $p=\ceu{p'\AtAnd{p''}}$.
    \begin{casex}
    \item$\lnot\isblocked(p',n)$.
      \begin{casex}
      \item $p'=\ceu{\Nop}$.
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop1}.
        Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
      \item\label{lem.x.det-nst.and-brk1}
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk1}.
        Hence~$p_1=p_2=\ceu{\clear(p'');\Break}$ and~$e_1=e_2=\nil$.
      \item\label{lem.x.det-nst.and-adv1}$p'\ne\ceu{\Nop},\ceu{\Break}$.
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv1}.  Thus
        there are derivations~$\pi_1'$ and~$\pi_2'$ such that
          \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
          \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
        for some~$p_1'$, $p_2'$, $e_1'$, $e_2'$.  By the induction
        hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
        Hence~$p_1=\ceu{{p_1'}\And{p''}}=\ceu{{p_2'}\And{p''}}=p_2$
        and~$e_1=e_1'=e_2'=e_2$.
      \end{casex}
    \item$\isblocked(p',n)$.
      \begin{casex}
      \item $p''=\ceu{\Nop}$.
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop2}.
        Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
      \item\label{lem.x.det-nst.and-brk2}$p''=\ceu{\Break}$.
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk2}.
        Hence~$p_1=p_2=\ceu{\clear(p');\Break}$ and~$e_1=e_2=\nil$.
      \item\label{lem.x.det-nst.and-adv2}$p''\ne\ceu{\Nop},\ceu{\Break}$.
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv2}.  Thus
        there are derivations~$\pi_1''$ and~$\pi_2''$ such that
          \pi_1''&\Vdash\<p'',n,\nil>\nst\<p_1'',n,e_1''>\\
          \pi_2''&\Vdash\<p'',n,\nil>\nst\<p_2'',n,e_2''>
        for some~$p_1''$, $p_2''$, $e_1''$, and~$e_2''$.  By the induction
        hypothesis, $p_1''=p_2''$ and~$e_1''=e_2''$.
        Hence~$p_1=\ceu{{p'}\And{p_1''}}=\ceu{{p'}\And{p_2''}}=p_2$
        and~$e_1=e_1''=e_2''=e_2$.
      \end{casex}
    \end{casex}
  \item$p=\ceu{p'\Or{p''}}$.
    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-expd}.
    Hence~$p_1=p_2=\ceu{{p'}\AtOr{(\CanRun(n);\,p'')}}$
    and~$e_1=e_2=\nil$.
  \item$p=\ceu{p'\AtOr{p''}}$.
    \begin{casex}
    \item$\lnot\isblocked(p',n)$.
      \begin{casex}
      \item $p'=\ceu{\Nop}$.
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-nop1}.
        Hence~$p_1=p_2=\clear(p'')$ and~$e_1=e_2=\nil$.
      \item $p'=\ceu{\Break}$.
        Similar to Case~\ref{lem.x.det-nst.and-brk1}.
      \item$p'\ne\ceu{\Nop},\ceu{\Break}$.
        Similar to Case~\ref{lem.x.det-nst.and-adv1}.
      \end{casex}
    \item$\isblocked(p',n)$.
      \begin{casex}
      \item $p''=\ceu{\Nop}$.
        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-nop1}.
        Hence~$p_1=p_2=\clear(p')$ and~$e_1=e_2=\nil$.
      \item $p''=\ceu{\Break}$.
        Similar to Case~\ref{lem.x.det-nst.and-brk2}.
      \item $p''\ne\ceu{\Nop},\ceu{\Break}$.
        Similar to Case~\ref{lem.x.det-nst.and-adv2}.
      \end{casex}
    \end{casex}
  \end{casex}
\end{proof}

\thmxdet*
\begin{proof}
  By induction on~$i$.  The theorem is trivially true if~$i=0$ and follows
  directly from Lemmas~\ref{lem.x.det-out} and~\ref{lem.x.det-nst} for~$i=1$.
  %%
  Suppose
  \[
    \delta\trans[1]\delta_1'\trans[i-1]\delta_1
    \quad\text{and}\quad
    \delta\trans[1]\delta_2'\trans[i-1]\delta_2\,,
  \]
  for some~$i>1$, $\delta_1'$ and~$\delta_2'$.
  There are two possibilities.
  \begin{casex}
  \item $\delta\out[1]\delta_1'$ and~$\delta\out[1]\delta_2'$.  Then, by
    Lemma~\ref{lem.x.det-out}, $\delta_1'=\delta_2'$, and by the induction
    hypothesis, $\delta_1=\delta_2$.
    %%
  \item $\delta\nst[1]\delta_1'$ and~$\delta\nst[1]\delta_2'$.  Then, by
    Lemma~\ref{lem.x.det-nst}, $\delta_1'=\delta_2'$, and by the induction
    hypothesis, $\delta_1=\delta_2$.\qedhere
  \end{casex}
\end{proof}

@@

  \item Similar to Case~\eqref{lem.props-nst-i.e}.\qedhere
  \end{enumerate:a}
\end{proof}

\begin{assumption}[Syntactic restrictions]\strut
  \label{ass.syn-rest}
  %%
  \begin{enumerate:a}
  \item\label{ass.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contain no
    occurrences of~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$,
    or~$\ceu{\Fin}$ expressions.  Consequently, for any~$n$,
    \[
      \<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>\,.
    \]
    %%
  \item\label{ass.syn-rest.loop} If~$p=\ceu{\Loop{p_1}}$ then all execution
    paths of~$p_1$ contain at least one occurrence of a
    matching~$\ceu{\Break}$ or an~$\ceu{\AwaitExt}$ expression.
    Consequently, for all~$n$, there are~$p_1'$ and~$e$ such that
    \[
      \<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>\,,
    \]
    where~$p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
  \end{enumerate:a}
\end{assumption}

@@

\begin{lemma}\strut
  \label{lem.rank-out}
  %%
  \begin{enumerate:a}
  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
    then~$\rank(\delta)=\rank(\delta')$.
  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
    then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate:a}
\end{lemma}
\begin{proof}\strut
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$,
  $\rank(\delta)=\<i,j>$, and~$\rank(\delta')=\<i',j'>$.
  \begin{enumerate:a}
  \item Suppose~$\<p,n,e>\outpush\<p',n',e'>$.  Then, by~\R{push},
    $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
    Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
    as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.
    %%
    It remains to be shown that~$i=i'$:
    \begin{align*}
      i&=\pot(p,e)
         \tag*{by Definition~\ref{def.rank}}\\
       &=\pot'(\bcast(p,e))
         \tag*{by Definition~\ref{def.pot}}\\
       &=\pot'(p')
         \tag*{since~$p'=\bcast(p,e)$}\\
       &=\pot'(\bcast(p',\nil))
         \tag*{by definition of~$\bcast$}\\
       &=\pot'(\bcast(p',e'))
         \tag*{since~$e'=\nil$}\\
       &=\pot(p',e')
         \tag*{by Definition~\ref{def.pot}}\\
       &=i'
         \tag*{by Definition~\ref{def.rank}}
    \end{align*}
    Therefore, $\<i,j>=\<i',j'>$.
